---
title: "Simulating hematopoietic phylogeny for SCM validation"
author: "Michael Tassia"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "",
                      fig.align = 'center',
                      out.width = "75%")
```

# Overview

The objective of this notebook is to document our validation strategy for applying stochastic character mapping as a technique for reconstructing somatic genotypes given a phylogenetic tree, substitution model, and a data set of observed genotype states at the tips (see [*Bollback 2006*](https://link.springer.com/article/10.1186/1471-2105-7-88) and [*Revell 2024*](https://peerj.com/articles/16505/)). As a ground-truth somatic phylogeny data set does not yet exist, we will leverage simulation as a framework for validation.

This notebook will first describe the approach, key functions, and expected outcomes. Afterwards, additional layers of technical and biological variation will be introduced, including:

- Sample size

- Sequencing depth

- Missing data

- ISM-violation rate

# Simulating hematopoiesis

We use the publicly available [`rsimpop` package](https://github.com/nangalialab/rsimpop) developed by Nick Williams at the Sanger Institute to simulate hematopoiesis under a birth-death model. This package implements the Gillepsie algorithm (see [here](https://lewiscoleblog.com/gillespie-algorithm) for basic overview) to simulate cellular compartments under a distribution of user-defined parameters and is described in the supplementary information of [*Mitchell et al. 2022*](https://www.nature.com/articles/s41586-022-04786-y#Sec32).

Here, we use `rsimpop` with parameters set following empirical estimates from the literature. Fitness coefficients manifest as increased symmetric division rates, and polygenic effects are assumed to be additive.

## Distribution of fitness effects (DFE)

`rsimpop` encodes fitness as increased rates of symmetric division (i.e., birth) for clones bearing a driver mutation. Here, we draw fitness effects from a gamma distribution using the parameters described in [Extended Data Fig 12 of *Mitchell et al. 2022*](https://www.nature.com/articles/s41586-022-04786-y/figures/17).

```{r}

# DFE function
genGammaFitness <- function(shape, rate) {
  function() rgamma(n = 1, shape = shape, rate = rate)
}
fitnessFn <- genGammaFitness(shape = 0.47, rate = 34) # parameters from Mitchell et al. 2022

# Plot DFE
hist(sapply(1:1e4, function(x) exp(fitnessFn()) - 1),
     breaks = 100, xlim = c(0, 1),
     xlab = "Selection coefficient", main = "")

```

## Simulate HSC population

Using the DFE above, we can simulate hematopoiesis with stochastic introduction of driver mutations using the `run_driver_process_sim()` wrapper in `rsimpop`. This function has several important parameters which will be defined below:

- `simpop = NULL`: Default setting; do not use a pre-existing simpop object

- `initial_division_rate = 0.1`: Default setting; parameter estimate for HSCs during development; only effects early population growth phase.

- `final_division_rate = 1/365`: Default setting; parameter estimate for HSCs during adulthood (concordant with the ABC estimates in [*Mitchell et al. 2022*](https://www.nature.com/articles/s41586-022-04786-y)).

- `target_pop_size = 1e5`: Default setting; mean parameter estimate for HSC population size at adulthood (concordant with the ABC estimates in [*Mitchell et al. 2022*](https://www.nature.com/articles/s41586-022-04786-y)).

- `nyears = 50`: User-defined age at collecting data. Validation will integrate across a distribution of ages.

- `fitness = fitnessFn`: User-defined fitness function defined above

- `drivers_per_cell_per_day = 2.0e-3 / 365`: User-defined count of driver mutations entering the population per day. This is the mean value estimated by ABC in [*Mitchell et al. 2022*](https://www.nature.com/articles/s41586-022-04786-y).

```{r}
SEED <- 1212121
initSimPop(SEED, bForce = TRUE)

sim <- run_driver_process_sim(
  simpop                   = NULL,  
  initial_division_rate    = 0.1,  
  final_division_rate      = 1/365,   
  target_pop_size          = 1e5, 
  nyears                   = 80,  
  fitness                  = fitnessFn, 
  drivers_per_cell_per_day = 2.0e-3 / 365  
)
```
We can then draw a sample from the HSC compartment and rescale the phylogeny from divisions to mutations.
```{r}

# Sample 100 cells from the population
st <- get_subsampled_tree(sim, 100)

# Re-scale the phylogeny by mutations
st_mut <- get_elapsed_time_tree(st, 
                                mutrateperdivision = 0, 
                                backgroundrate = 16.8/365)

# Plot driver events on the phylogeny
plot_tree_events(st_mut,
                 cex.label = 0,
                 legpos = NULL)

```